Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison with Guidelines for Reporting

### 2023-08-31 paper.qmd

-   Despite calls to use moving averages or digital filter, binned averages remain the most common data processing strategy to determine V̇O~2max~.
-   Different processing strategies lead to varying V̇O~2max~ values at the median as well as at the individual level.
-   Researchers should at minimum report their processing strategy in detail, or preferably share their raw oxygen uptake data and analysis code.

@@

A total of N = 72 exercise tests were analyzed for this study. Due to a miscalculation, the preregistration had incorrectly stated a number of 76 tests. The data were from previous research on the metabolic profile of endurance runners [@quittmann_accepted; @schwarz2022]. The tested individuals were experienced distance runners (15 female, 54 male; three of the males participated in both studies). The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann_accepted] and March to October 2021 [@schwarz2022], respectively, while using identical exercise protocols and test equipment. Participants run on a treadmill (saturn 300/100, h/p/cosmos sports & medical 127 GmbH, Nussdorf-Traunstein, Germany) with 1% inclination for 8 minutes at a velocity of 2.8 m·s^-1^ as a warm-up. After preparing the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m·s^-1^ for two minutes and subsequently increased velocity by 0.15 m·s^-1^ every 30 seconds. The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.

@@

The calculation intervals for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @fig-duration). 30-second intervals were most common to define V̇O~2max~, while some authors also often employed shorter (10-20 s) and longer (60 s) periods.

```{r}
#| echo: false
#| fig-cap: "Total durations of the calculation interval of V̇O~2max~ in the reviewed studies. STILL UP TO DEBATE WHETHER RATHER PLOT (VERTICAL OR HORIZONTAL BARS?) OR TABLE."
#| label: fig-duration
#| out-width: '90%'

knitr::include_graphics("./plots/duration_count.png")
```

@@

Breath-based averages seem to be more common (\~8%) than reported previously (\<1%) [@midgley2007], but less common than assessed in self-reporting (\~17%) [@robergs2010]. The increasing proportion of breath-based averages may be explained by publications in the recent years advocating for their use [@robergs2003; @robergs2010]. The length of the calculation interval for averaging is highly diverse within the literature (see @fig-duration). This may reflect contradictory recommendations [@howley1995; @robergs2003]. There is no consensus on using shorter or longer calculation intervals, and it is unclear whether an optimal interval duration even exists. As different interval durations influence V̇O~2max~ by more than 5% with respect to median levels (see @fig-comparison), the exact reporting of the data processing strategy remains essential for interpretation.

@@

The different data processing strategies found in the literature directly bias the V̇O~2max~ determination (see @fig-comparison), and as such influence the classification of individuals, the evaluation of training success, and the assessment of V̇O~2max~ attainment. In accordance with previous findings [@johnson1998; @astorino2009; @sell2021] longer calculation intervals lead to lower V̇O~2max~ values (see @fig-comparison). The analyzed data shows median differences up to 5-7% between processing strategies, which is in accordance with previous research [@martin-rincon2019]. Some studies reported even higher mean differences of up to 20% [@myers1990], but only when including raw breath-by-breath data in the comparison. The evaluation of unprocessed raw data for its maximum is highly erroneous and as such is not performed in research (see @fig-duration); so there is no reason to compare it to other strategies. While previous research was often conducted in sedentary or recreationally trained individuals, we now present evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in trained athletes. Since training interventions in trained athletes typically show improvements in V̇O~2max~ in the range of 0-6% [@parmar2021; @midgley2007a], variation caused by differing data processing strategies is in the same range and can bias the evaluation of their success.

@@

The exact impact of data processing strategies on the V̇O~2max~ is highly individual. Most research did present only comparisons of mean values, with results in accordance with those found here [@martin-rincon2019]. Data processing strategies may impact V̇O~2max~ values with varying magnitudes at the individual level. For example, for 10% of the investigated athletes a binned time average of 5 seconds leads to a V̇O~2max~ \<3% higher than by a 30-second average, while for another 10% of the investigated athletes the V̇O~2max~ was \>6% higher (see @fig-comparison). Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparison of group results for meta-analyses or group classifications. However, at the individual level these equations can only be applied with a large margin of error. Changes in the impact of different data processing strategies on V̇O~2max~ range from 1-2% in some individuals to more than 10% in others. Hence when evaluating V̇O~2max~ data from different tests in a single individual obtained by using different processing methods, there is no way to accurately compare these values even when the processing strategies are reported. While the comparison of V̇O~2max~ from different processing strategies require their reporting for an sufficient analysis on a group level, the raw data from each test is required on the individual level.

@@

To compare and evaluate V̇O~2max~ values from different studies, knowledge of the underlying data processing strategies is crucial. Our review demonstrates that almost half of the studies measuring V̇O~2max~ did not describe their processing strategy. Other aspects of the data processing, such as outlier filtering or the rationale for the chosen procedure were only in rare instances reported (see @tbl-reporting). @tbl-recommendation lists seven items that should be reported to provide sufficient information on the data processing strategy used to determine V̇O~2max~. These items may be reported in form of a checklist, as an in-text enumeration or in sentence form. An example paragraph containing all the relevant information for the original data presented in this paper [@quittmann_accepted; @schwarz2022] would be:

@@

DISCUSS IF WE SHOULD CHANGE THE ORDER OF RECOMMENDATIONS/GUIDELINES AND LIMITATIONS.

>>>>

-   Despite calls to use moving averages or digital filter, binned averages remain the most common data processing strategy to determine V̇O~2max~.
-   Different processing strategies lead to varying V̇O~2max~ values at the median as well as at the individual level.
-   Researchers are advised to report their processing strategy in detail, or preferably share their raw oxygen uptake data and analysis code.

@@

A total of N = 72 exercise tests were analyzed for this study. Due to a miscalculation, the preregistration had incorrectly stated a number of 76 tests. The data were from previous research on the metabolic profile of endurance runners [@quittmann2022; @schwarz2022]. The tested individuals were experienced distance runners (15 female, 54 male; three of the males participated in both studies). The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann2022] and March to October 2021 [@schwarz2022], respectively, while using identical exercise protocols and test equipment. Participants run on a treadmill (saturn 300/100, h/p/cosmos sports & medical 127 GmbH, Nussdorf-Traunstein, Germany) with 1% inclination for 8 minutes at a velocity of 2.8 m·s^-1^ as a warm-up. After preparing the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m·s^-1^ for two minutes and subsequently increased velocity by 0.15 m·s^-1^ every 30 seconds. The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.

@@

The calculation intervals for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @tbl-duration). 30-second intervals were most common to define V̇O~2max~, while some authors also often employed shorter (10-20 s) and longer (60 s) periods.

```{r}
#| echo: false
#| tbl-cap: "Total durations of the calculation interval of V̇O~2max~ in the reviewed studies."
#| label: tbl-duration

load("./data/intervals.Rda")

i_df <- as.data.frame(table(i_data$param_duration))

knitr::kable(
  i_df,
  col.names = c("Interval duration (s)", "Count"),
  align = "r"
)
```

@@

Breath-based averages seem to be more common (\~8%) than reported previously (\<1%) [@midgley2007], but less common than assessed in self-reporting (\~17%) [@robergs2010]. The increasing proportion of breath-based averages may be explained by publications in the recent years advocating for their use [@robergs2003; @robergs2010]. The length of the calculation interval for averaging is highly diverse within the literature (see @tbl-duration). This may reflect contradictory recommendations [@howley1995; @robergs2003]. There is no consensus on using shorter or longer calculation intervals, and it is unclear whether an optimal interval duration even exists. As different interval durations influence V̇O~2max~ by more than 5% with respect to median levels (see @fig-comparison), the exact reporting of the data processing strategy remains essential for interpretation.

@@

The different data processing strategies found in the literature directly bias the V̇O~2max~ determination (see @fig-comparison), and as such influence the classification of individuals, the evaluation of training success, and the assessment of V̇O~2max~ attainment. In accordance with previous findings [@johnson1998; @astorino2009; @sell2021] longer calculation intervals lead to lower V̇O~2max~ values (see @fig-comparison). The analyzed data shows median differences up to 5-7% between processing strategies, which is in accordance with previous research [@martin-rincon2019]. Some studies reported even higher mean differences of up to 20% [@myers1990], but only when including raw breath-by-breath data in the comparison. The evaluation of unprocessed raw data for its maximum is highly erroneous and as such is not performed in research (see @tbl-duration); so there is no reason to compare it to other strategies. While previous research was often conducted in sedentary or recreationally trained individuals, we now present evidence that a similar effect of data processing strategies on V̇O~2max~ exists in trained athletes.

Since training interventions in trained athletes typically show improvements in V̇O~2max~ in the range of 0-6% [@parmar2021; @midgley2007a], variation caused by differing data processing strategies is in the same range and can bias the evaluation of their success. Together with biological [@katch1982] and technical [@winkert2021] variability, data processing is just one of several sources of variation in the process of V̇O~2max~ determination, but it is one that can easily be controlled without the need for multiple testing.

@@

The exact impact of data processing strategies on the V̇O~2max~ is highly individual. Most research did present only comparisons of mean values, with results in accordance with those found here [@martin-rincon2019]. Data processing strategies may impact V̇O~2max~ values with varying magnitudes at the individual level. For example, for 10% of the investigated athletes a binned time average of 5 seconds leads to a V̇O~2max~ \<3% higher than by a 30-second average, while for another 10% of the investigated athletes the V̇O~2max~ was \>6% higher (see @fig-comparison). Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparison of group results for meta-analyses or group classifications. However, at the individual level these equations can only be applied with a large margin of error. Changes in the impact of different data processing strategies on V̇O~2max~ range from 1-2% in some individuals to more than 10% in others. Hence when evaluating V̇O~2max~ data from different tests in a single individual obtained by using different processing methods, there is no way to accurately compare these values even when the processing strategies are reported. While the comparison of V̇O~2max~ from different processing strategies require their reporting for a sufficient analysis on a group level, the raw data from each test is required on the individual level.

@@

To compare and evaluate V̇O~2max~ values from different studies, knowledge of the underlying data processing strategies is crucial. Our review demonstrates that almost half of the studies measuring V̇O~2max~ did not describe their processing strategy. Other aspects of the data processing, such as outlier filtering or the rationale for the chosen procedure were only in rare instances reported (see @tbl-reporting). @tbl-recommendation lists seven items that should be reported to provide sufficient information on the data processing strategy used to determine V̇O~2max~. These items may be reported in form of a checklist, as an in-text enumeration or in sentence form. An example paragraph containing all the relevant information for the original data presented in this paper [@quittmann2022; @schwarz2022] would be:

@@

# Summary

Despite calls for standardization, researchers use a variety of data processing strategies to determine the V̇O~2max~ from raw gas exchange data. Based on current reporting practices we developed a checklist that can serve as a guideline for reporting data processing methods for V̇O~2max~ determination. We advise authors to follow the reporting guidelines and ideally share anonymized raw data to improve the reproducibility of research in exercise physiology.

### 2023-08-31 scripts/review.Rmd
Adjust to next feedback round

# Show data processing type for breath-by-breath measurements only
table(ext$proc_combination[ext$type == "bbb"], useNA = "ifany")

@@

knitr::include_graphics("../plots/duration_count.png")
```

>>>>

# Show data processing type for breath-by-breath measurements only
ext$total_comb <- paste(ext$proc_combination, ext$parameters, sep = "_")
table(ext$total_comb[ext$type == "bbb"], useNA = "ifany")

## Show processing type X interval combinations

table(ext$proc_combination[ext$type == "bbb"], useNA = "ifany")

@@

knitr::include_graphics("../plots/duration_count.png")

# save data for table generation as alternative to a plot
i_data <- ext[ext$ptype == "time" & !is.na(ext$ptype), ]
save(i_data, file = "../data/intervals.Rda")
```
